Modal Verification Patterns for Systems Software

### 2025-09-13 paper.tex
+

Most of this coordination comes through reasoning about what \emph{will be true after local operations are published to memory},
or what \emph{will be true after remote operations become visible locally}~\cite{sieczkowski2015separation,dang2022compass,fsl}. Each of these flavors of contingent reasoning
depend on use of certain \emph{fence} or \emph{memory barrier} operations, which block execution of a CPU until either all incoming or outgoing
memory transactions are complete.
%\section{Resources in Systems Software}
%\label{sec:systemsoft}
%\subsection{Speculation in Consistency} \todo{speak about the weak-mem modalities in here.}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}
\subsubsection{Actor Model} An alternative to deal with the shared mmemory concurrency complication, i.e., data races, is using actors ~\cite{actormodel} which appears as a way of sequentializing the concurrent interleaved execution with some dependency on the \emph{knowledge} about the other actors.

\section{Contingency Decomposition of a System}
In this section, we argue the position that \emph{modalities}  
should be a go-to approach when specifying and verifying low-level systems code.
We explain how the concept of a \emph{resource context} helps guide the design of new modalities
for verification of systems code, and
we justify our perspective by discussing prior systems that have used modalities for systems verification
successfully, arguing that they fit into the verification design pattern we articulate,
and explaining how this approach might apply to other systems verification challenges.

To explain our ideas in the general systems understanding, we briefly recap some of the background and themes our ideas build on again, casting them in a certain way to bring out the relevance of our philosophy.
\subsection{Encoding Modalities in Program Logics}
Program logics which would like to enjoy modal propositions need to have an underlying model that allows encoding of the propositions. Since . Kripke model is the model all program logics directly or indirectly, sits on. Kripke model can be roughly seen as a triple of set of worlds (e.g., program states), a binary relation between worlds (e.g., transition relations between states) and a state interpretation function.  
\paragraph{Encoding Hoare Triples} With Pratt's original representation, we can encode where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$. However, this modal representation requires to be given a semantics which is in terms of an underlying model.

Conventionally, a Kripke model with its constituent must be picked. Gordon give a Kripke model for an actor-based hybrid modal logic ~\cite{gordon2019modal} in which the transition relation is indexed by a particular command and a particular actor which relates pre- and post-states of a particular actor referenced by a nominal executing the command. Gordon encodes this hybrid logic in Dafny program logic ~\cite{leino2010dafny} with the help of two state invariants and rely-guarantee assertions checking the stability of invariants. Wagner et.al. takes the similar approach for the design of their separation logic and pick up a Kripke model for reference counted memory reasoning principles.

>>>>

Most of this coordination comes through reasoning about what \emph{will be true after local operations are published to memory},
or what \emph{will be true after remote operations become visible locally}~\cite{sieczkowski2015separation,dang2022compass,fsl}. Each of these flavors of contingent reasoning
depend on use of certain \emph{fence} or \emph{memory barrier} operations, which block execution of a CPU until either all incoming or outgoing
memory transactions are complete.
%\section{Resources in Systems Software}
%\label{sec:systemsoft}
%\subsection{Speculation in Consistency} \todo{speak about the weak-mem modalities in here.}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}
\subsubsection{Actor Model} An alternative approach to addressing the shared memory concurrency complication, i.e., data races, is to use actors ~\cite{actormodel} by eliminating shared mutable state and enforcing updates on exclusive local mutable state via exchanging immutable messages among processes. This exhibits itself as a way of sequentializing the concurrent interleaved execution with some dependency on the \emph{knowledge} about other actors. In a typical consensus algorithm, an operation is committed if nodes reach an agreement on it.

Therefore, as expected, the specification of an actor in the current view would refer to the other actors' states and has to validate that the shared knowledge amongst actors is preserved ~\cite{gordon2019modal}.
\section{Contingency Decomposition of a System}
In this section, we argue the position that \emph{modalities}  
should be a go-to approach when specifying and verifying low-level systems code.
We explain how the concept of a \emph{resource context} helps guide the design of new modalities
for verification of systems code, and
we justify our perspective by discussing prior systems that have used modalities for systems verification
successfully, arguing that they fit into the verification design pattern we articulate,
and explaining how this approach might apply to other systems verification challenges.

To explain our ideas in the general systems understanding, we briefly recap some of the background and themes our ideas build on again, casting them in a certain way to bring out the relevance of our philosophy.
\subsection{Encoding Modalities in Program Logics}
Program logics which would like to enjoy modal propositions need to have an underlying model that allows encoding of the propositions. Since . Kripke model is the model all program logics directly or indirectly, sits on. Kripke model can be roughly seen as a triple of set of worlds (e.g., program states), a binary relation between worlds (e.g., transition relations between states) and a state interpretation function.  
\paragraph{Encoding Hoare Triples}  With Pratt's original representation, we can encode where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$. However, this modal representation requires a semantics, which is defined in terms of an underlying model.

Conventionally, a Kripke model with its constituent must be picked. Gordon provides a Kripke model for an actor-based hybrid modal logic ~\cite{gordon2019modal}, in which the transition relation is indexed by a particular command and a particular actor, relating pre- and post-states of a specific actor referenced by a nominal executing the command. Gordon encodes this hybrid logic in Dafny program logic ~\cite{leino2010dafny} with the help of two state invariants and rely-guarantee assertions checking the stability of invariants. Wagner et al. take a similar approach to the design of their separation logic and utilize a Kripke model for reference-counted memory reasoning principles.
